Unifying intraspecific and interspecific variation in tropical tree mortality

### 2016-10-19 ms/ms.Rnw
Format for supplementary materials
- layout for title etc
- add figures
- add references to figure in main text
Edits to manuscript and figure code

% Enable cross referencing to Evidence document
\usepackage{xr}
\externaldocument{ms-suppinfo}
\title{Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework}

@@

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{Add something significant here. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse libero sem, pharetra posuere mauris eget, tincidunt auctor ex. Sed a magna eget lorem posuere euismod ac in enim. Nam porttitor placerat est, quis finibus quam efficitur nec. Pellentesque velit diam, scelerisque accumsan iaculis dignissim, vulputate vitae lorem. Maecenas eros sapien, semper quis velit nec, porta convallis est. Vestibulum porttitor diam ac massa porta pharetra. In a viverra justo. Ut aliquam purus non neque bibendum, ultricies tincidunt mi tempus. Aenean dignissim sollicitudin pharetra. Integer aliquet odio odio, vel bibendum mi lobortis nec. Aliquam non fermentum elit, id pellentesque ex.}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits, and temporal plot-scale effects; these distinct drivers are yet to be integrated within a cohesive and predictive framework. Here we apply classic survival analysis techniques within a cross-validated hierarchical Bayesian framework to derive instantaneous mortality functions that integrate individual, species, and temporal effects for 203 tropical rain forest tree species at Barro Colorado Island (BCI) Panama. We show that mortality models that separate growth-dependent and growth-independent components, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that the effects of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability in species mortality rates, and that temporal variation in mortality was large.  Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects, needed to effectively predict tree mortality.
\end{abstract}

@@

\dropcap{R}ates of plant mortality are known to be highly variable among individuals, among coexisting species, between forests and from year-to-year [refs]. This variation has considerable consequences for forest structure and dynamics. Death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools\cite{Chave:2005hz}, while at the same time opening up a gap that restarts a successional race, during which 10's-100's more plants may die while competing for a spot in the sun. In models of forest dynamics, variation in mortality rates has been found to have a larger impact on forest structure than absolute growth rates\cite{Johnson-2016}. As such, improving our ability to understand the mortality process is therefore a priority for making skilful predictions about forest dynamics and carbon, water or nutrient cycles; especially in an era of rapid environmental change.

Two substantial challenges exist when studying tree mortality in tropical rainforests. The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates (Poorter et al 2008). This requirement arises in part because the observable outcome -- alive vs. dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate). In classic survival analysis \cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$\cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
\end{equation}
Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required. As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys in long-term monitoring plots containining thousands of individuals and spanning several decades\cite{Anderson-Teixeira-2015a}.

The other major challenge when studying mortality -- similar to that faced in other fields like medicine and engineering -- is to derive suitable shapes for the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality. Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate\cite{Kobe:1995tw,Wyckoff:2002ul}, traits\cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size\cite{Vieilledent:2010fv, Hurst:2011ky}. Generally, however, these influences have not been formalised as hazard functions\cite{Zens:2003jv}. Yet such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function\cite{Hawkes:2000ib;Zens:2003jv,McDowell:2011dr}. This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.

A particular challenge is to integrate the observed conflicting relationships between mortality and growth observed within and across species. Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies provide broad support, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. Yet across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality than individuals from slower growing species\cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume). Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack\cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events\cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

@@

In this study, we reconcile both intraspecific and interspecific drivers of instantaneous mortality by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework. To derive this unified model, we compare the predictive accuracy of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig. \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. The simplest model (Fig. \ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (Fig. \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases.

Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached. The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models. Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (Moorcroft 2001; McDowell et al 2011). However, these alternative models have not previously been systematically compared.

In deriving the unified model, we address the following specific questions:
\begin{enumerate}\itemsep 0pt
  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
  \item What is the proportion of inter-specific variability explained by wood density?
  \item What is the proportion of total variability explained by species, census, baseline hazard and growth-dependent hazard?
  \item Does mortality vary substantially between censuses?
\end{enumerate}
Answering these questions involves comparing predictive accuracy of alternative models. To achieve this, we used K-fold cross-validation (Fig. \ref{fig1}D). Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations. For example, some of our models took over a two weeks to complete. We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increasing model complexity (see methods).

@@

\subsection*{Mortality strongly varies over time}
Tree mortality progressively decreased over time. Between 1995 and 2010, the highest proportion of trees death occured between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and lastly 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a significant increase in predictive accuracy (i.e. lower logarithmic loss; Fig. \ref{fig2}A). Assuming the long-term average census effect ($\delta$) is 1, then our model predicted deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.

@@

@@

Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions. In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of observed growth rates were negative. While there are physiological processes (e.g. water depletion and bark drop) that can lead to trunk shrinkages overtime\cite{Pastur:2007kl}, it is biologically impossible for growth (i.e. the creation of new biomass) to be negative\cite{Clark:2007ve}. To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e. observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors\cite{Condit:2012nz}. This dataset consisted of \Sexpr{nobs_meas_err_data(BCI_dbh_error_data)} randomly selected trees that had their diameter measured twice within a 30 day period\cite{Condit:2012nz}. Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error. The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma(BCI_dbh_error_data)}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter. Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual (SUPP).

@@

For all models we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}. We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic\cite{Gelman:1992ts, Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior.

\subsection*{Model evaluation}
Predictive accuracy was quantified by estimating the average log loss across K-folds for held-out data, $\bar{\mathcal{L}}$ (Fig. \ref{fig1}). Logarithmic loss -- commonly known as log loss, and indicated by $\mathcal{L}$-- measures the accuracy of a model by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome (Fig ??). By minimizing the $\bar{\mathcal{L}}$, we maximize predictive skill. The average log loss across all individuals for the $k$th fold of held-out data, $\mathcal{L}_k$, is given by
\begin{equation} \label{eq:logloss}
\mathcal{L}_k = - \frac1{N_k}\,\sum_{i=1}^{N_k}\big(S_i\,\log(p_{i,t_2\rightarrow t_3}) + ( 1 - S_i)\,\log(1 - p_{i,t_2\rightarrow t_3})\big),
\end{equation}
where $N_k$ refers to the number of observations in fold.

@@

\acknow{
J.C was supported by the Science and Industry Endowment Fund (SIEF; RP04-174). M.W was supported by a fellowship from the Australian Research Council (ARC). D.F. was supported by an ARC discovery grant. We thank Lachlan McCalman and Daniel Steinberg for statistical advice; Anias Gibert etc etc.  Fig. 1a uses images by Tracey Saxby, IAN Image Library (http://ian.umces.edu/imagelibrary)}

>>>>

% Enable cross referencing to Evidence document
\usepackage{xr}
\externaldocument{supp}

\title{Unifying intra- and inter-specific variation in tropical tree mortality: towards a mechanistic framework}

@@

% Please add here a significance statement to explain the relevance of your work
% 120 words
\significancestatement{Tree mortality is a fundamental demographic process which affects forest dynamics and consequently, a terrestrial ecosystems ability to store carbon. Here, for the first time, we use observed survival records over a 15 year period from more than 180,000 individuals, to simultaneously estimate growth-dependent and growth-independent (i.e. death by stocastic chance) mortality across 203 tropical forest tree species. We found that growth-dependent mortality was the predominate factor influencing tree mortality rates at Barro Colorado Island (BCI). Furthermore, we found that wood density most strongly influenced mortality rates by decreasing growth-independent mortality. However, wood density only accounted for a small fraction of the overall species variability observed in growth-independent mortality rates, suggesting that there must be other species traits that strongly affect mortality. Which traits and how many of them, however, remains unknown.}

@@

% 250 words
\begin{abstract}
Tree death is a fundamental process driving population dynamics, nutrient cycling, and evolution within plant communities. While past research has identified factors influencing tree mortality across a variety of scales -- including an individual's past growth, species-level functional traits and temporal plot-scale effects -- these distinct drivers are yet to be integrated within a cohesive and predictive framework. Here we apply classic survival analysis techniques within a cross-validated hierarchical Bayesian framework to derive instantaneous mortality functions that integrate individual, species, and temporal effects for 203 tropical rain forest tree species at Barro Colorado Island (BCI) Panama. We show that mortality models that partition mortality into growth-dependent and growth-independent components, use stem diameter growth rather than basal-area growth, and attribute the effect of wood density to growth-independent death, outperform alternative formulations. Moreover, we show that the effects of wood density -- a prominent trait known to influence tree mortality -- explains only a small proportion of the total variability in species mortality rates, and that temporal variation in mortality was large. Our analysis also suggests that growth-dependent processes -- and thus competition -- predominate as the biggest driver of mortality at BCI. Combined, this study provides a framework for predicting individual-level mortality in highly diverse tropical forests, but also highlights how little we now about the causes of species-level and temporal plot-scale effects needed to effectively predict tree mortality.
\end{abstract}

@@

\dropcap{R}ates of plant mortality are known to be highly variable among individuals, among coexisting species, between forests and from year-to-year [refs]. This variation has considerable consequences for forest structure and dynamics. For example, death of a single large tree can transfer up to 20000 kg of carbon from live-to-decaying carbon pools\cite{Chave:2005hz}. Furthermore it creates a gap in the canopy that can restart a successional race, during which 10's-100's more plants may die while competing for a spot in the sun. In models of forest dynamics, variation in mortality rates have been found to have a larger impact on forest structure than variation in absolute growth rates\cite{Johnson-2016}. As such, improving our ability to understand the mortality process is a priority for making accruate predictions about forest and carbon dynamics as well as water or nutrient cycles; especially in an era of rapid environmental change.

Two difficulties arise when studying tree mortality in tropical rainforests. The first is that large population sizes and long periods of observation are required to make inferences into how various factors affect mortality rates\cite{Poorter:2008iu}. This requirement arises in part because the observable outcome -- alive vs. dead -- is one-step removed from the variable we ideally want to measure: $\lambda_i(t)$, the instantaneous rate of mortality for individual $i$ at time $t$ (also called a hazard rate). In classic survival analysis\cite{Kleinbaum:2012tn}, the probability that an individual dies between times $t_2$ and $t_3$ is expressed as a function of $\lambda_i(t)$:
\begin{equation} \label{eq:mortality_model}
p_{i,t_2\rightarrow t_3} = 1 - \exp\left(- \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t\right).
\end{equation}
Here $\int_{t_2}^{t_3} \lambda_i(t) {\rm d}t$ is the ``cumulative hazard'' between $t_2$ and $t_3$\cite{Kleinbaum:2012tn}. The observed survival outcome $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) is then a realisation of this probability
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}).
\end{equation}
Since trees are generally long-lived, and because we are trying to estimate the shape of a continuous function from binary data, large sample-sizes are required. As such, detailed studies of forest tree mortality have only recently become possible, with the accumulation of growth and survival data from repeat surveys in long-term monitoring plots containining thousands of individuals and spanning several decades\cite{Anderson-Teixeira-2015a}.

The other difficulty when studying mortality -- similar to that faced in other fields like medicine and engineering -- is determining the shape of the underlying instantaneous mortality (hazard) function that accurately reflect observed patterns in mortality. Past research has identified a range of factors with significant effects on tree survival, including an individual's growth rate\cite{Kobe:1995tw,Wyckoff:2002ul}, traits\cite{Poorter:2008iu,Kraft:2010kq,Wright:2010fl,AubryKientz:2013dg}, and size\cite{Vieilledent:2010fv, Hurst:2011ky}. Generally, however, these influences have not been formalised as hazard functions\cite{Zens:2003jv}. Yet, such functions are critical for forecasting temporal patterns in tree mortality, and consequently, changes in ecosystem function\cite{Hawkes:2000ib;Zens:2003jv,McDowell:2011dr}. This is because hazard functions explicitly define the instantaneous rate of death per unit time and consequently provide vital information on when a death is likely to occur and how various covariates affect this.

A particular challenge in developing hazard functions for plants is to integrate the observed conflicting relationships between mortality and growth within and across species. Within species, mortality rates tend to decrease for faster growing individuals, presumably because those individuals have superior carbon budgets and are thus better able to tolerate or repair diverse stresses\cite{Kobe:1995tw, Hawkes:2000ib, Keane:2001db}. Empirical studies provide broad support of this theory, with many highlighting non-linear declines in mortality with increased growth\cite{Kobe:1997vy, Wyckoff:2002ul,Vieilledent:2010fv}. By contrast, across species, there tends to be a strong trade-off between growth and survival, with individuals from faster growing species having higher mortality rates than individuals from slower growing species\cite{Poorter:2008iu, Wright:2010fl}. This trade-off may arise via traits with antagonistic effects, such as wood density (biomass per unit wood volume). Denser wood -- which is more expensive to build and thus slows growth rates -- conveys enhanced physical strength, reduced stem breakage\cite{Putz:1983wu,vanGelder:2006exa,Chave:2009iy}, embolism\cite{Hacke:2001kj, Jacobsen:2005fx}, fungal or pathogen attack\cite{Augspurger:1984wx}, and ultimately reduced mortality due to stochastic events\cite{Poorter:2008iu, Chave:2009iy, Kraft:2010kq, Wright:2010fl}.

Another major challenge is using binary observations (e.g. alive = 0; dead = 1) to infer the contribution of growth-dependent and growth-independent (i.e. stochastic events) hazards on tree mortality rates. By partitioning mortality rates into these two processes model predictive capacity may be improved and it allows us to narrow in on whether growth-dependent or growth-independent processes are the predominate cause of death in an ecosystem. For example, are tree deaths in rainforest predominately a result of competitive influences (i.e. growth-dependent processes) or are they more likely to result from growth-independent events such as treefall? More importantly, however, such partitions will allow us to examine how plant traits affect each hazard. For example, higher wood density is thought to convey higher survival by decreasing mortality aociated with stoachastic events such as those listed above. However, research has also shown that wood density may increase tolerance to low growth rates, and thus alter the growth-dependent mortality. For example, studies have found that species with high wood density are more tolerant to both shade \cite{vanGelder:2006exa}, neighbor competition \cite{Kunstler:2016km} and drought induced carbon starvation \cite{McDowell:2011dr}. Consequently, wood density may affect a trees response to both growth-dependent and growth-independent hazards. The question is, which hazard does it affect most?

@@

In this study, we reconcile both intraspecific and interspecific drivers and attempt to partition instantaneous mortality rates into growth-dependent and growth-independent hazards by developing a unified model that incorporates individual-level, species-level, and temporal effects, within a cross-validated Bayesian framework. To derive this unified model, we compare the predictive accuracy of three alternative cumulative hazard functions ($\lambda_i(t)$; Fig. \ref{fig1}B) in predicting patterns of tropical tree mortality in 203 tropical rainforest trees at Barro Colorado Island (BCI), Panama. The simplest model (Fig. \ref{fig1}B, left), assumes a single baseline hazard, which is constant irrespective of a plant's growth rate. The second model (Fig. \ref{fig1}B, middle) describes a growth-dependent hazard which assumes the risk of dying eventually declines to zero as growth increases. Following past work \cite[e.g.][]{Kobe:1995tw, Wyckoff:2000un, Kunstler:2005bn, Vieilledent:2010fv}, we assume the growth-mortality relationship is adequately captured by a negative exponential curve, with mortality decaying exponentially until an asymptote at zero is reached. The third model (eq. Fig. \ref{fig1}B, right) is the summation of the two previous models. Specifically, this model allows $\lambda_i(t)$ to decrease exponentially with increasing growth rate, but, unlike a standard negative exponential, will asymptote at some baseline hazard instead of zero. The advantage of this model is that it effectively accounts for two independent causes of mortality: death due to growth and death due to chance (i.e. stochastic events such as wind throw and fire). Combined, these three models capture a variety of functional responses previously proposed, including effects represented in current vegetation models (Moorcroft 2001; McDowell et al 2011). However, these alternative models have not previously been systematically compared.

In deriving the unified model, we address the following specific questions:
\begin{enumerate}\itemsep 0pt
  \item Does a mortality function including both growth-dependent and independent terms outperform a model including only one of these?
  \item Which growth measure (diameter increment growth or area growth) best captures growth-dependent mortality?
  \item Does including wood density improve model predictive capacity, and if so, does it predominately affect baseline hazard, growth-dependent hazard or both?
  \item What is the proportion of inter-specific variability explained by wood density?
  \item What is the proportion of total variability explained by species, census, baseline hazard and growth-dependent hazard?
  \item Does mortality rates vary substantially between censuses?
\end{enumerate}
Answering these questions involves comparing predictive accuracy of alternative models. To achieve this, we used K-fold cross-validation (Fig. \ref{fig1}D). Cross-validation is a typical approach to model evaluation, and assesses a model’s accuracy when extrapolating to data not used in model calibration, thereby indicating the model’s true predictive accuracy \cite{Hooten:2015, Stone:1997tg}. Evaluating models in this way is computationally expensive, making it virtually impossible to run all combinations. For example, some of our models took over a two weeks to complete. We therefore fit models across five iterative stages of model fitting and comparison, corresponding to the six questions evaluated above, and gradually increasing model complexity (see methods).

@@

\subsection*{Mortality over time}
Tree mortality progressively decreased over time. Between 1995 and 2010, the highest proportion of trees death occured between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 2)}\%) and lastly 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_unique, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a significant increase in predictive accuracy (i.e. lower logarithmic loss; Fig. \ref{fig2}A). Assuming the long-term average census effect ($\delta$) is 1, then our model predicted deaths were \Sexpr{census_effects(final_model,1)}\% higher in 1995\textendash 2000, and \Sexpr{census_effects(final_model,2)}\% and \Sexpr{census_effects(final_model,3)}\% lower in 2000\textendash 2005 and 2005\textendash 2010, respectively.

@@

Example referencing suppmat figure (Fig. \ref{figS1}).

@@

Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions. In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of observed growth rates were negative. While there are physiological processes (e.g. water depletion and bark drop) that can lead to trunk shrinkages overtime\cite{Pastur:2007kl}, it is biologically impossible for growth (i.e. the creation of new biomass) to be negative\cite{Clark:2007ve}. To ensure our mortality model was not biased by these implausible values we first developed a model to estimate \textit{``true growth''}(i.e. observed growth minus observation error). Details of this model can be found in supplementary material (REF, FigS1).

@@

For all models we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}\cite{StanDevelopmentTeam:2015uz}. We executed three independent chains and in all cases modelled parameters converged within 2000 iterations. Convergence was assessed through both visual inspection of chains and reference to the Brooks-Gelman-Rubin convergence diagnostic\cite{Gelman:1992ts, Brooks:1998ju}. After discarding the first 2000 iterations as ‘burn in’, a further 2000 iterations were taken from the joint posterior. These models were then cross validated using 10-fold cross validation (REF).

\subsection*{Model evaluation}
Predictive accuracy was quantified by estimating the average log loss across 10-folds for held-out data, $\bar{\mathcal{L}}$ (Fig. \ref{fig1}). Logarithmic loss -- commonly known as log loss, and indicated by $\mathcal{L}$-- measures the accuracy of a model by penalizing incorrect predictions based on how wrong the predicted probability is from the observed outcome (Fig. \ref{fig1}C). By minimizing the $\bar{\mathcal{L}}$, we maximize predictive skill. The average log loss across all individuals for the $k$th fold of held-out data, $\mathcal{L}_k$, is given by
\begin{equation} \label{eq:logloss}
\mathcal{L}_k = - \frac1{N_k}\,\sum_{i=1}^{N_k}\big(S_i\,\log(p_{i,t_2\rightarrow t_3}) + ( 1 - S_i)\,\log(1 - p_{i,t_2\rightarrow t_3})\big),
\end{equation}
where $N_k$ refers to the number of observations in fold.

\subsection*{Posthoc correlations}

@@

\acknow{
J.C was supported by the Science and Industry Endowment Fund (SIEF; RP04-174). M.W was supported by a fellowship from the Australian Research Council (ARC). D.F. was supported by an ARC discovery grant. We thank Rich FitzJohn, Lachlan McCalman and Daniel Steinberg for computational advice; Anias Gibert etc etc.  Fig. 1a uses images by Tracey Saxby, IAN Image Library (http://ian.umces.edu/imagelibrary)}

### 2016-10-19 ms/supp.Rnw
Edits to manuscript and figure code

We did this by first modelling the observed initial ($D_{i,1}$) and final diameter ($D_{i,2}$) of each individual, $i$, as random realisations from a normal distribution centerd on their respective true estimates (i.e. $\widehat{D_{i,1}}$ and $\widehat{D_{i,2}}$), with the standard deviation fixed at \Sexpr{obs_error_sigma} (i.e. the observed measurement error):

@@

\subsection{Mortality model}
Our observations are for survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$. We model these observations as random realizations from a Bernoulli distribution
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}),
\end{equation}
where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$ *(eq. \ref{eq:mortality_model}). As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
\begin{equation}
  \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t \approx \lambda_i(t_2) (t_3-t_2).
\end{equation}

@@

\clearpage
\section{Appendix 2 - Example stan code}

Include example model here

Describe how to get other models?

@@

\begin{figure}[!hb]
\centering
\includegraphics[width=15cm,height=20cm,keepaspectratio]{figures/supp_fig1}
\caption{ About this figure}
\label{figS1}
\end{figure}

\clearpage
\begin{figure}[!hb]
\centering
\includegraphics[width=15cm,height=20cm,keepaspectratio]{figures/supp_fig2}
\caption{ About this figure}
\label{figS2}
\end{figure}

>>>>

Field measurements of diameter inevitably include observation error, errors that may ultimately bias conclusions. In our dataset, \Sexpr{prop_neg_growth(BCI_model_dataset_true_unique)}\% of observed growth rates were negative. While there are physiological processes (e.g. water depletion and bark drop) that can lead to trunk shrinkages overtime\cite{Pastur:2007kl}, it is biologically impossible for growth (i.e. the creation of new biomass) to be negative\cite{Clark:2007ve}. To ensure our model was not biased by these implausible values we estimated \textit{``true growth''}(i.e. observed growth minus observation error) by first using another BCI dataset collected explicitly to examine observation errors\cite{Condit:2012nz}. This dataset consisted of \Sexpr{nobs_meas_err_data(BCI_dbh_error_data)} randomly selected trees that had their diameter measured twice within a 30 day period\cite{Condit:2012nz}. Assuming no growth occurred over this period, any discrepancy in measurements can be attributable to observation error. The standard deviation of these discrepancies was estimated to be \Sexpr{obs_error_sigma(BCI_dbh_error_data)}, meaning that 95\% of the measurement errors were within $\pm$ 1.47 cm of the true diameter. Having an estimate of observation error, we then used Bayesian point estimation to maximize the joint posterior of the \textit{``true''} initial and final diameter, and subsequently, \textit{``true growth''} for each individual.

We modelled true growth by first estimating the observed initial ($D_{i,1}$) and final diameter ($D_{i,2}$) of each individual, $i$, as random realisations from a normal distribution centerd on their respective true estimates (i.e. $\widehat{D_{i,1}}$ and $\widehat{D_{i,2}}$), with the standard deviation fixed at \Sexpr{obs_error_sigma(BCI_dbh_error_data)} (i.e. the observed measurement error):

@@

\subsection{Mortality model}
We modelled tree mortality using survival outcomes $S_{i, t_2\rightarrow t_3}$ (0 = alive, 1 = died) for each individual $i$ between censuses at time $t_2$ and $t_3$. We model these observations as random realizations from a Bernoulli distribution
\begin{equation}
S_{i,t_2\rightarrow t_3} \sim \textrm{Bernoulli}(p_{i,t_2\rightarrow t_3}),
\end{equation}
where $p_{i,t_2\rightarrow t_3}$ is the probability individual $i$ dies in the interval $t_2$--$t_3$ *(eq. \ref{eq:mortality_model}). As we only have a point estimate of $\lambda_i(t)$ at $t=t_2$, the cumulative hazard from  $t_2$--$t_3$ is approximated as
\begin{equation}
  \int_{t_2}^{t_3} \lambda_i(t) {\rm d}t \approx \lambda_i(t_2) (t_3-t_2).
\end{equation}

@@

\clearpage
\section{Appendix 2 - Stan code for final model fit with species random effects}

<<echo=FALSE>>=
fullmodelcode()
@

@@

\begin{figure}[!hb]
\centering
\includegraphics{figures/supp_fig1.png}
\caption{Observed vs predicted individual growth rates and stem diameters}
\label{figS1}
\end{figure}

\clearpage
\begin{figure}[!hb]
\centering
\includegraphics{figures/supp_fig2}
\caption{Map of BCI 50 ha plot showing tree recruits (red crosses) against gap index for censuses 1985\textendash 1990 and 1990\textendash 1995. Colour gradient indicates the minimum amount of canopy shading above 2 m in a 5 year census. Dark colours indicate dense canopy shading, light colours indicate little or no canopy shading above 2 m.}
\label{figS2}
\end{figure}

\clearpage
\begin{figure}[!hb]
\centering
\includegraphics{figures/supp_fig3}
\caption{alpha}
\label{figS3}
\end{figure}

\clearpage
\begin{figure}[!hb]
\centering
\includegraphics{figures/supp_fig4}
\caption{alpha}
\label{figS4}
\end{figure}

\clearpage
\begin{figure}[!hb]
\centering
\includegraphics{figures/supp_fig5}
\caption{alpha}
\label{figS5}
\end{figure}
